Interim report\\ Measles risk assessment, modelling and cost analysis\\ \vspace{2 mm} {\large David T S Hayman, Tim Carpenter,\\ Jonathan C Marshall, Mick Roberts, Nigel P French}

### 2014-07-04 interimreport1.Rnw
Draft final interim report
Bibliography complete, compiles, etc.

\usepackage{times}
\usepackage{hyperref}
\usepackage{tikz}
\usepackage{bibunits}
\usepackage{natbib}
\usepackage{graphics}
\usepackage{amsmath}
\usepackage{indentfirst}
\usepackage[utf8]{inputenc}
\usepackage{graphicx}
\usepackage{todonotes}

@@

\title{Interim report\\ Measles risk assessment, modeling and cost analysis\\ \vspace{2 mm} {\large David T S Hayman, Tim Carpenter,\\ Jonathan C Marshall, Mick Roberts, Nigel P French}}
\author{mEpiLab and EpiCentre,\\ Infectious Diseases Research Centre,\\
Massey University,\\
Palmerston North 4442,\\
New Zealand\\
\href{mailto: D.T.S.Hayman@massey.ac.nz}{D.T.S.Hayman@massey.ac.nz}}  %\texttt formats the text to a typewriter style font
\date{\today}  %\today is replaced with the current date
\maketitle

@@

New Zealand has been working towards elimination of endemic (domestic) measles virus transmission, but has suffered from small, but significant outbreaks of measles after measles introductions from abroad. In this interim report we reviewed the draft \emph {Progress Towards Measles Elimination in New Zealand - Final} report from the New Zealand Ministry of Health to the World Health Organization (WHO) Western Pacific Region. We identified additional analyses that may help understand risk of infection in New Zealand, began analyses of the cost of the current 2013--2014 outbreak and began modeling measles outbreaks to understand how they may be controlled, and to evaluate the likelihood that measles may become endemic once more. Our key findings were:
\begin{itemize}
\item The \emph {Progress Towards Measles Elimination in New Zealand - Final} report was of high quality and contained substantial information and useful analyses.
\item New Zealand is at risk of frequent measles importation due to travel and endemic measles elsewhere in the globe.
\item The cost of the current 2013--2014 measles outbreak is estimated to be at least \$750,000.
\item Analyses of outbreak data suggest that measles basic reproductive number (Rv, the number of secondary infections) values often include 1 and this year, 2014, is well above one. This analyses suggest improved vaccination is a requisite to prevent measles becoming endemic again.
\end{itemize}

@@

The report concludes that New Zealand's surveillance system has been performing well and that the Ministry is confident that measles has not been circulating since June 2012 and has not become endemic in NZ. We agree with the statement that measles did not become endemic and provide some preliminary analyses on the outbreaks since endemic measles elimination (see section \ref{sec:epidemic_modeling}) that gives information regarding the likelihood of measles persisting within the population and becoming endemic, including important analysis of the current outbreak.

We agree with the report conclusions that testing for measles is performed appropriately within the required timeframe. Clearly improving inter-laboratory communication and collaboration and timeliness of the testing and reporting is necessary for rapid responses to measles introductions following measles control. Vaccination coverage presented in the report and to ourselves confirms that immunisation levels are approaching 94\% for MMR dose one (birth cohorts 2009 and 2010) and 89\% for MMR dose two (birth cohorts 2006 and 2007). However, only Asian and Pacific ethnicities have consistently had MMR dose one coverage approaching or exceeding 95\% for cohorts from 2007 onwards, and thus we agree with the report's conclusions that timeliness and coverage of vaccination need improving. This is particularly in light of our preliminary modeling and risk analyses results (sections \ref{sec:risk_analyses} and \ref{sec:epidemic_modeling}).

The analyses we believe could further inform the understanding of risk from measles infection are:
\begin{itemize}
\item Multivariate modelling to account for confounding within the univariate analyses.\todo{one L or two in modelling?}
\item Inclusion of multivariate model results with local level immunity data to develop a risk map for New Zealand.
\item Analyses of changing risk factors through time during outbreaks.
\item Modelling transmission chains in the population to understand effective reproduction number.
\item Development of risk maps to understand measles importation.
\end{itemize}

@@

For our initial analyses, we use arrivals data from New Zealand immigration (\href{http://www.immigration.govt.nz/}{www.immigration.govt.nz}) as a proxy for human movement to and from New Zealand. We collated country population size, measles incidence and measles vaccination cover from the WHO (\href{http://www.who.int/research/en/}{www.who.int/research/en/}).\todo{this sentence seems to need a bit more after it? Something like "to produce pretty maps"}

<<echo=FALSE>>=
library(xtable)
dpdf<-read.csv("mapdata.csv",header=T)
# output top countries
topimmigration <- dpdf[order(-dpdf$immigration),] 
immigration<-topimmigration[1:10,c(1,4)]
topincidence <- dpdf[order(-dpdf$incidence),] 
incidence<-topincidence[1:10,c(1,2)]
incidence[,2]<-round(incidence[,2],0)
topvaccine <- dpdf[order(dpdf$cover),] 
vaccinecover<-topvaccine[1:10,c(1,3)]
toprisk <- dpdf[order(-dpdf$risk),] 
risk<-(toprisk[1:10,c(1,5)])
risk[,2]<-round(risk[,2],0)
@

@@

\begin{table}
\caption{Risk of measles importation to New Zealand in 2012}
<<results=tex,echo=FALSE>>=
# This chunk uses "results=tex"
library(Hmisc)
latex(risk, file="", table.env=FALSE,rowname=NULL)
@
\label{table:risk12}
\end{table}\todo{these tables need country rather than id as the first column name}

@@

\subsection{Future risk analyses}\sec{risk_analyses}
We received the raw EpiSurv measles case data from The Institute of Environmental Science and Research Ltd (ESR) on 27 June 2014. Initial analyses of those data (not shown) suggest that we require denominator data to perform multivariate analyses to avoid confounding results due to a lack of independence among risk factors. Specifically for the multivariate analyses we wish to perform that detect interactions we require $\textsf{Age} \times \textsf{Prioritised Ethnicity} \times \textsf{NZDep}$ data for New Zealand to test whether interactions among case covariates provide additional information on risk over the univariate analyses performed in the report reviewed above. These data only became available to us on 3 July 2014, provided by the University of Otago.

The following analyses are in progress, for inclusion in later reports:
\begin{itemize}
\item Multivariate regression analyses to assess interactions between risk factors that may confound the univariate analyses.
\item Update of the importation risk analyses using a broader range of years, including modeling the trend in importations.
\end{itemize}
Additional data we believe would enable us and the Ministry to better understand measles risk is fine scale (lower than District Health Board (DHB)) immunisation coverage data. Thus the data gap that we have that will hinder us providing fine scale risk maps are:
\begin{itemize}
\item Meshblock (or census area unit) level immunisation coverage data to allow targeted immunisation and understanding of risk at a fine scale level.
\item The number of New Zealanders arriving from abroad each year, the countries to which they travelled, and length of travel. \todo{added a bit here you may want to check}
\end{itemize}

@@

From 16 December, 2013 through 19 June, 2014 there were 201 confirmed measles cases in New Zealand (note 14 of these occurred before Jan 2014, so 187 occurred from Jan 2013 -- 19 June 2014). The number of cases by age group is shown in Table~\ref{table:freq}. Of these 201 cases, 34 (17\%) were admitted to hospital with the highest proportion occurring in the youngest (< 15 months) and oldest (> 19 years) age groups, 0.47 and 0.33, respectively.\todo{make these percentages?}

@@

Regression analyses showed statistically significant associations between cost of hospitalisation and three variables, length of hospitalisation, case age and year of case, and a less strong association with case gender (Table~\ref{table:regression}). Results showed an increase of approximately \$358 per day of hospitalisation, \$139 for a female case, and \$1 per year of case diagnosis. The cost was found to decrease by approximately \$5 per year of case age.\todo{later on we could assess whether or not a linear term for case age is appropriate, or what if any interaction there is between age and length of stay?}

@@

\subsection{Cost analysis discussion}
The results presented here are based on available data. While some of the data are complete and detailed, this is not true of all the data. In order to perform an accurate analysis of the current measles outbreak in New Zealand, more complete data are needed. For instance, age, gender, ethnicity, year of discharge, length of stay and estimated cost data are available for cases reported by publicly funded hospitals. In addition to this information, similar data would be needed for cases occurring outside the period 2011--2013 at publicly funded hospitals. In addition, similar data would be needed for non-publicly funded hospitals, e.g. private clinics.

@@

In other outbreaks, the average cost per measles case was estimated to be US\$ 254, US\$ 276, and US\$ 307 for Canada, the Netherlands, and the UK, respectively (Carabin, et al., 2002). This and other findings will be compared and contrasted with New Zealand costs, once more complete New Zealand data are made available.
The containment of a single case (also 2 secondary cases) of measles in 2004 in Iowa, USA was estimated to cost US\$ 142,542. In this outbreak, more than 2500 hours of personnel time were needed to investigate and respond to the outbreak (Dayan et al., 2005). They estimated direct costs per case to be less than US\$ 500. 
The annual cost for long-term care of people with moderate of severe mental retardation over a period of 50 years is estimated at US\$ 31,059 and US\$ 78,448, respectively \cite{prouty0}. In 2000 expenditures for care in large state mental retardation/developmental disabilities (MR/DD) facilities continued to increase and reached a national annual average of US\$ 113,864 per person. In 2000 the average annual expenditures for care in large state MR/DD facilities were \$ 113,864. The cost of a case of measles was estimated to range from \$ 71 (no complications and no hospitalisation) to \$ 29,556 (encephalitis and hospitalisation for 8.7 days). They estimated the annual cost of measles in the US with its vaccination program to be \$1,234,083 (52.5\% direct cost and 47.5\% indirect cost) \cite{zhou4}.

@@

\subsection{Future cost-benefit analyses}
Using the results above we aim to:
\begin{itemize}
\item Estimate the costs for targeted vaccination, based either on the univariate analyses presented to date in the \emph {Progress Towards Measles Elimination in New Zealand - Final} report or adjusted if any additional risk groups are identified in the multivariate analyses or modeling.
\end{itemize}

@@

\section{Modeling measles epidemics}\label{sec:epidemic_modeling}

@@

The quantity that determines whether an epidemic will occur is the basic reproduction number of the infection, $R_0$\todo{not sure if you want it like this throughout?}. This is defined as the expected number of secondary infections that would arise from a single primary infection introduced into a fully susceptible population \cite{anderson91, diekmann0}. If $R_0 > 1$ an epidemic will occur following an introduction of infection. The best estimate we had for measles in New Zealand was R0 = 12.8 \cite{roberts4}. The basic reproduction number of the infection under vaccination, Rv\todo{or $R_v$ ?}, is the expected number of secondary infections that would arise from a single primary infection introduced into a vaccinated population at equilibrium and is a robust indicator of the performance of a vaccination schedule. If $R_v < 1$ epidemics are prevented. The case reproduction number of the infection at time $t$, $R_t$, is the expected number of secondary infections that arise from a single infection at a particular time and depends on the number in the population who are susceptible.

To understand the level of immunity in the population and how likely an outbreak was of becoming endemic, we estimated Rv from all the outbreaks since 2009. To do this we estimated Rt, following an adaptation of the methods in \cite{obidia12}\todo{there's 2 references in \texttt{?est.R0.TD}}, and used these to estimate the Rv of measles in the recent outbreaks, which helps understand the transmission dynamics of measles in the partially immune population.

To estimate Rt we were required to compute the generation time\todo{what is this? define?} for measles. We used a lognormal distribution with mean 12.0 and standard deviation (s.d.) 3.5 from \cite{klinkenberg11}. We then estimated Rt using the methods adapted from \cite{obidia12} from the incidence data for each outbreak. We defined outbreaks in the dataset given their temporal and geographic correlations (Figure~\ref{fig:dhbcases}). The outbreaks we used in out analyses are shown in Figure~\ref{fig:outbreaks}.\todo{this repeats a bit of the previous paragraph - could do with rewrite}

@@

\begin{figure}
     \centering
     \includegraphics[width=0.8\textwidth]{outbreaks_for_R0.pdf}
     \caption{Measles data classified as outbreaks for reproductive number of the infection (Rv) estimation}
     \label{fig:outbreaks}
\end{figure}

\begin{figure}
     \centering
     \includegraphics[width=0.8\textwidth]{averageR0.pdf}
     \caption{Estimates of Rv (R0) for the outbreaks each year, as classified in Figure~\ref{fig:outbreaks} (page~\pageref{fig:outbreaks})}
     \label{fig:r0}
\end{figure}

The estimated Rv for each outbreak is shown in Figure~\ref{fig:r0}. The probability density of the Rv estimates for each outbreak all include one. Of particular note is the ongoing outbreak, which has an Rv well above one and thus we may expect this outbreak to persist if conditions remain the same.\todo{to be fair, this is because it's ongoing - as we haven't seen a decline yet, $R_0$ is necessarily over one.}

These analyses also imply that the regular (approximately yearly) importation of measles is an ongoing process. Given the risk of importation of measles as highlighted in section~\ref{sec:risk_analyses} is likely to continue, these analyses suggest substantial efforts are required to maintain the level of immunisation to high enough levels that measles does not become endemic. The measles outbreak in 2011--2012 had an Rv of just greater than one, and yet it persisted for over 12 months. This implies that the current outbreak may persist within the population for a substantial period, given it's Rv is approximately twice that of the 2011-2012 outbreak. A caveat to this and other Rv estimates is that the 2013--2014 outbreak may include some sporadic cases and thus the true basic reproductive numbers may be lower than estimated. However, sub-clinical and underreporting may lower the estimate. The relative contributions of both to our estimates are currently unknown.

\subsection{Summary of modeling}

\begin{itemize}
\item Regular introductions of measles pose an ongoing threat to New Zealand's efforts to eliminate measles (also see section~\ref{sec:risk_analyses}).
\item The reproduction number for measles in a partially immune population is often close to one, suggesting increased population level immunity is required to prevent this measles persisting.
 \item The reproduction number for measles in the current outbreak is well over one, suggesting that this outbreak has the potential to become endemic.\todo{this seems a bit extreme -- caveat?}
\end{itemize}

\section{Future modeling}
Future modeling we aim to perform are:
\begin{itemize}
\item An update of previous ODE models of measles outbreak risk in the overall population according the differing vaccine coverage scenarios.
\item Model measles outbreaks with differing scenarios of measles importation into various population groups based on current introduction rates.
\end{itemize}

\section {Summary of key findings}
\begin{itemize}
\item New Zealand is at risk of frequent measles importation due to travel and endemic measles elsewhere in the world.
\item The cost of the current measles outbreak is estimated to be at least \$750,000.
\item Analyses of outbreak data suggest that measles Rv values often include 1 and in this year, 2014, are well above one. This analysis suggests improved vaccination is a requisite to prevent measles becoming endemic again.
\end{itemize}

@@

\bibitem[Carabin et al. (2003)]{carabin3}
Carabin, H., W.~J. Edmunds, M. Gyldmark, P. Beutels, D. Lévy-Bruhl, H. Salo, U.~K. and Griffiths (2003).
\newblock The cost of measles in industrialised countries.
\newblock \emph{Vaccine}, 21,4167--4177.

@@

\bibitem[Dayan et al.(2005)]{dayan5}
Dayan, G.~H, I.~R. Ortega-Sánchez, C.~W. LeBaron, M.~P. Quinlisk, and the Iowa Measles Response Team (2005).
\newblock The cost of containing one case of measles: the economic impact on the public health infrastructure - Iowa, 2004.
\newblock \emph{Pediatrics), 116:e1; DOI:10/1542/peds.2004-2512.

\bibitem[Diekmann et al.(2000)]{diekmann0}
Diekmann, O. &  J.~A.~P. Heesterbeek (2000).
\newblock \emph{Mathematical epidemiology of infectious diseases: model building, analysis and interpretation}.
Chichester: Wiley.

\bibitem[Edmunds et al.(2000)]{edmunds0}
Edmunds, W.~J., N.~J. Gay, M. Kretzschmar, R.~G. Pebody and H. Wachman (2000).
\newblock \The pre-vaccination epidemiology of measles, mumps and rubella in Europe: implications for modelling studies.
\newblock \emph{Epidemiology and Infection}, 125, 635--650.

\bibitem[Filia et al.(2007)]{filia7}
Filia, A., A. Brenna, A. Paná, G.~M. Cavallaro, M. Massari, and M. Atti, (2007).
\newblock Health burden and economic impact of measles-related hospitalization in Italy, 2002-2003.
\newblock \emph{BMC Public Health}, 7,169

@@

\bibitem[Geyer and Thompson(1995)]{gay98}
Gay, N.~J., L. Pelletier, and P. Duclos (1998).
\newblock Modelling the incidence of measles in Canada: an assessment of the options for vaccination policy.
\newblock \emph{Vaccine}, 16, 794--801.

@@

\bibitem[Koopmanschap(1998)][koopmanschap98}
Koopmanschap, M.~A. (1998).
\newblock Cost-of-illness studies: useful for health policy?
\newblock \emph{Pharmacoeonomics}, 14, 143--148.

@@

\bibitem[Obidia et al.(2012)]{obidia12}
Obadia, T., R. Haneef and P.~Y., Boëlle (2012).
\newblock The R0 package: a toolbox to estimate reproduction numbers for epidemic outbreaks.
\newblock \emph{BMC Medical Informatics and Decision Making}, 2012, 12--147.

\bibitem[Parker et al.(2006)]{parker6}
Parker, A.~A., W. Staggs, G.~H. Dayan, T.~R. Ortega-Sánchez, P.~A. Rota, L. Lowe, P. Boardman, R. Teclaw, C. Graves, and C.~W. LeBaron (2006).
\newblock Implications of a 2005 measles outbreak in Indiana for sustained elimination of measles in the United States.
\newblock \emph{The New England Journal of Medicine}, 355, 447--455.

@@

@@

\bibitem[WHO (2013)]{who13}
World Health Organisation measles media centre, January (2013)
\newblock Geneva: World Health Organization.
http://www.who.int/mediacentre/news/notes/2013/measles_20130117/en/, accessed July 1, 2014.

>>>>

\usepackage{times}
\usepackage{hyperref}
\usepackage{tikz}
\usepackage{bibunits}
%\usepackage{natbib}
\usepackage{graphics}
\usepackage{amsmath}
\usepackage{indentfirst}
\usepackage[utf8]{inputenc}
\usepackage{graphicx}
\usepackage{eurosym}
\usepackage{todonotes}

@@

\title{Interim report\\ Measles risk assessment, modelling and cost analysis\\ \vspace{2 mm} {\large David T S Hayman, Tim Carpenter,\\ Jonathan C Marshall, Mick Roberts, Nigel P French}}
\author{mEpiLab and EpiCentre,\\ Infectious Diseases Research Centre,\\
Massey University,\\
Palmerston North 4442,\\
New Zealand\\
\href{mailto: D.T.S.Hayman@massey.ac.nz}{D.T.S.Hayman@massey.ac.nz}}  %\texttt formats the text to a typewriter style font
\date{\today}  %\today is replaced with the current date
\maketitle

@@

New Zealand has been working towards elimination of endemic (domestic) measles virus transmission, but has suffered from small, but significant outbreaks of measles after measles introductions from abroad. In this interim report we reviewed the draft \emph {Progress Towards Measles Elimination in New Zealand - Final} report from the New Zealand Ministry of Health to the World Health Organization (WHO) Western Pacific Region. We identified additional analyses that may help understand risk of infection in New Zealand, began analyses of the cost of the current 2013--2014 outbreak and began modelling measles outbreaks to understand how they may be controlled, and to evaluate the likelihood that measles may become endemic once more. Our key findings were:
\begin{itemize}
\item The \emph {Progress Towards Measles Elimination in New Zealand - Final} report was of high quality and contained substantial information and useful analyses.
\item New Zealand is at risk of frequent measles importation due to travel and endemic measles elsewhere in the globe.
\item The cost of the current 2013--2014 measles outbreak is estimated to be at least \$750,000.
\item Analyses of outbreak data suggest that measles basic reproductive number ($R_v$, the number of secondary infections) values often include 1 and this year, 2014, is well above one. This analyses suggest improved vaccination is a requisite to prevent measles becoming endemic again.
\end{itemize}

@@

The report concludes that New Zealand's surveillance system has been performing well and that the Ministry is confident that measles has not been circulating since June 2012 and has not become endemic in NZ. We agree with the statement that measles did not become endemic and provide some preliminary analyses on the outbreaks since endemic measles elimination (see section \ref{sec:epidemic_modelling}) that gives information regarding the likelihood of measles persisting within the population and becoming endemic, including important analysis of the current outbreak.

We agree with the report conclusions that testing for measles is performed appropriately within the required timeframe. Clearly improving inter-laboratory communication and collaboration and timeliness of the testing and reporting is necessary for rapid responses to measles introductions following measles control. Vaccination coverage presented in the report and to ourselves confirms that immunisation levels are approaching 94\% for MMR dose one (birth cohorts 2009 and 2010) and 89\% for MMR dose two (birth cohorts 2006 and 2007). However, only Asian and Pacific ethnicities have consistently had MMR dose one coverage approaching or exceeding 95\% for cohorts from 2007 onwards, and thus we agree with the report's conclusions that timeliness and coverage of vaccination need improving. This is particularly in light of our preliminary modelling and risk analyses results (sections \ref{sec:risk_analyses} and \ref{sec:epidemic_modelling}).

The analyses we believe could further inform the understanding of risk from measles infection are:
\begin{itemize}
\item Multivariate modelling to account for confounding within the univariate analyses.
\item Inclusion of multivariate model results with local level immunity data to develop a risk map for New Zealand.
\item Analyses of changing risk factors through time during outbreaks.
\item Modelling transmission chains in the population to understand effective reproduction number.
\item Development of risk maps to understand measles importation.
\end{itemize}

@@

For our initial analyses, we use arrivals data from New Zealand immigration (\href{http://www.immigration.govt.nz/}{www.immigration.govt.nz}) as a proxy for human movement to and from New Zealand. We collated country population size, measles incidence and measles vaccination cover from the WHO (\href{http://www.who.int/research/en/}{www.who.int/research/en/}). Note this uses all immigration of foreign nationals, coming for whatever purpose. We used the WHO data to determine per capita measles cases for each year and used this and the number of immigrants to New Zealand to begin to understand where measles was likely to be imported from.

<<echo=FALSE>>=
library(xtable)
dpdf<-read.csv("mapdata.csv",header=T)
# output top countries
colnames(dpdf)[1]<-c("country")
topimmigration <- dpdf[order(-dpdf$immigration),] 
immigration<-topimmigration[1:10,c(1,4)]
topincidence <- dpdf[order(-dpdf$incidence),] 
incidence<-topincidence[1:10,c(1,2)]
incidence[,2]<-round(incidence[,2],0)
topvaccine <- dpdf[order(dpdf$cover),] 
vaccinecover<-topvaccine[1:10,c(1,3)]
toprisk <- dpdf[order(-dpdf$risk),] 
risk<-(toprisk[1:10,c(1,5)])
risk[,2]<-round(risk[,2],0)
@

@@

\begin{table}
\caption{Risk of measles importation to New Zealand in 2012}
<<results=tex,echo=FALSE>>=
# This chunk uses "results=tex"
library(Hmisc)
latex(risk, file="", table.env=FALSE,rowname=NULL)
@
\label{table:risk12}
\end{table}

@@

\subsection{Future risk analyses}\label{sec:risk_analyses}
We received the raw EpiSurv measles case data from The Institute of Environmental Science and Research Ltd (ESR) on 27 June 2014. Initial analyses of those data (not shown) suggest that we require denominator data to perform multivariate analyses to avoid confounding results due to a lack of independence among risk factors. Specifically for the multivariate analyses we wish to perform that detect interactions we require $\textsf{Age} \times \textsf{Prioritised Ethnicity} \times \textsf{NZDep}$ data for New Zealand to test whether interactions among case covariates provide additional information on risk over the univariate analyses performed in the report reviewed above. These data only became available to us on 3 July 2014, provided by the University of Otago.

The following analyses are in progress, for inclusion in later reports:
\begin{itemize}
\item Multivariate regression analyses to assess interactions between risk factors that may confound the univariate analyses.
\item Update of the importation risk analyses using a broader range of years, including modelling the trend in importations.
\end{itemize}
Additional data we believe would enable us and the Ministry to better understand measles risk is fine scale (lower than District Health Board (DHB)) immunisation coverage data. Thus the data gap that we have that will hinder us providing fine scale risk maps are:
\begin{itemize}
\item Meshblock (or census area unit) level immunisation coverage data to allow targeted immunisation and understanding of risk at a fine scale level.
\item The number of New Zealanders arriving from abroad each year, the countries to which they travelled, and length of travel.
\end{itemize}

@@

From 16 December, 2013 through 19 June, 2014 there were 201 confirmed measles cases in New Zealand (note 14 of these occurred before Jan 2014, so 187 occurred from Jan 2013 -- 19 June 2014). The number of cases by age group is shown in Table~\ref{table:freq}. Of these 201 cases, 34 (17\%) were admitted to hospital with the highest proportion occurring in the youngest (< 15 months) and oldest (> 19 years) age groups, 47\% and 33\%, respectively.

@@

Regression analyses showed statistically significant associations between cost of hospitalisation and three variables, length of hospitalisation, case age and year of case, and a less strong association with case gender (Table~\ref{table:regression}). Results showed an increase of approximately \$358 per day of hospitalisation, \$139 for a female case, and \$1 per year of case diagnosis. The cost was found to decrease by approximately \$5 per year of case age.

@@

\subsection{Cost analysis discussion}
The results presented here are based on available data. While some of the data are complete and detailed, this is not true of all the data. In order to perform an accurate analysis of the current measles outbreak in New Zealand, more complete data are needed. For instance, age, gender, ethnicity, year of discharge, length of stay and estimated cost data are available for cases reported by publicly funded hospitals. In addition to this information, similar data would be needed for cases occurring outside the period 2011--2013 at publicly funded hospitals. In addition, similar data would be needed for non-publicly funded hospitals, e.g. private clinics. Other factors that we aim to investigate are whether or not a linear term for case age is appropriate, or what if any interaction there might be between age and length of stay in hospital.

@@

In other outbreaks, the average cost per measles case was estimated to be US\$ 254, US\$ 276, and US\$ 307 for Canada, the Netherlands, and the UK, respectively (Carabin, et al., 2002). This and other findings will be compared and contrasted with New Zealand costs, once more complete New Zealand data are made available.
The containment of a single case (also 2 secondary cases) of measles in 2004 in Iowa, USA was estimated to cost US\$ 142,542. In this outbreak, more than 2500 hours of personnel time were needed to investigate and respond to the outbreak (Dayan et al., 2005). They estimated direct costs per case to be less than US\$ 500. 
The annual cost for long-term care of people with moderate of severe mental retardation over a period of 50 years is estimated at US\$ 31,059 and US\$ 78,448, respectively \cite{prouty1}. In 2000 expenditures for care in large state mental retardation/developmental disabilities (MR/DD) facilities continued to increase and reached a national annual average of US\$ 113,864 per person. In 2000 the average annual expenditures for care in large state MR/DD facilities were \$ 113,864. The cost of a case of measles was estimated to range from \$ 71 (no complications and no hospitalisation) to \$ 29,556 (encephalitis and hospitalisation for 8.7 days). They estimated the annual cost of measles in the US with its vaccination program to be \$1,234,083 (52.5\% direct cost and 47.5\% indirect cost) \cite{zhou4}.

@@

\subsection{Future cost-benefit analyses}
Using the results above we aim to:
\begin{itemize}
\item Estimate the costs for targeted vaccination, based either on the univariate analyses presented to date in the \emph {Progress Towards Measles Elimination in New Zealand - Final} report or adjusted if any additional risk groups are identified in the multivariate analyses or modelling.
\end{itemize}

@@

\section{Modelling measles epidemics}\label{sec:epidemic_modelling}

@@

The quantity that determines whether an epidemic will occur is the basic reproduction number of the infection, $R_0$. This is defined as the expected number of secondary infections that would arise from a single primary infection introduced into a fully susceptible population \cite{anderson91, diekmann0}. If $R_0 > 1$ an epidemic will occur following an introduction of infection. The best estimate we had for measles in New Zealand was $R_0$ = 12.8 \cite{roberts4}. The basic reproduction number of the infection under vaccination, $R_v$, is the expected number of secondary infections that would arise from a single primary infection introduced into a vaccinated population at equilibrium and is a robust indicator of the performance of a vaccination schedule. If $R_v < 1$ epidemics are prevented. The case reproduction number of the infection at time $t$, $R_t$, is the expected number of secondary infections that arise from a single infection at a particular time and depends on the number in the population who are susceptible.

To understand the level of immunity in the population, the transmission dynamics of measles in the partially immune population and how likely an outbreak was of becoming endemic, we estimated $R_v$ from all the outbreaks since 2009. To do this we estimated $R_t$, following an adaptation of the methods in \cite{obidia12,wallinga4}. We were required to compute the generation time for measles to do so. The generation time is the average time an index case infects others after becoming infected. We used a lognormal distribution with mean 12.0 and standard deviation (s.d.) 3.5 from \cite{klinkenberg11}. We then estimated $R_t$ from the incidence data for each outbreak, defining outbreaks in the dataset given their temporal and geographic correlations (Figure~\ref{fig:dhbcases}).The outbreaks we used in out analyses are shown in Figure~\ref{fig:outbreaks}.

@@

\begin{figure}
     \centering
     \includegraphics[width=0.8\textwidth]{outbreaks_for_R0.pdf}
     \caption{Measles data classified as outbreaks for reproductive number of the infection ($R_v$) estimation}
     \label{fig:outbreaks}
\end{figure}

\begin{figure}
     \centering
     \includegraphics[width=0.8\textwidth]{averageR0.pdf}
     \caption{Estimates of $R_v$ ($R_0$) for the outbreaks each year, as classified in Figure~\ref{fig:outbreaks} (page~\pageref{fig:outbreaks})}
     \label{fig:r0}
\end{figure}

The estimated $R_v$ for each outbreak is shown in Figure~\ref{fig:r0}. The probability density of the $R_v$ estimates for each outbreak all include one. Of particular note is the ongoing outbreak, which has an $R_v$ well above one and thus we may expect this outbreak to persist if conditions remain the same.An important caveat to this analysis is that because this is an ongoing outbreak, and not in decline, $R_0$ is necessarily over one.

These analyses also imply that the regular (approximately yearly) importation of measles is an ongoing process. Given the risk of importation of measles as highlighted in section~\ref{sec:risk_analyses} is likely to continue, these analyses suggest substantial efforts are required to maintain the level of immunisation to high enough levels that measles does not become endemic. The measles outbreak in 2011--2012 had an $R_v$ of just greater than one, and yet it persisted for over 12 months. This implies that the current outbreak may persist within the population for a substantial period, given it's $R_v$ is approximately twice that of the 2011-2012 outbreak. A caveat to this and other $R_v$ estimates is that the 2013--2014 outbreak may include some sporadic cases and thus the true basic reproductive numbers may be lower than estimated. However, sub-clinical and underreporting may lower the estimate. The relative contributions of both to our estimates are currently unknown.

\subsection{Summary of modelling}

\begin{itemize}
\item Regular introductions of measles pose an ongoing threat to New Zealand's efforts to eliminate measles (also see section~\ref{sec:risk_analyses}).
\item The reproduction number for measles in a partially immune population is often close to one, suggesting increased population level immunity is required to prevent this measles persisting.
 \item The reproduction number, $R_v$, for measles in the current outbreak is well over one, suggesting that this outbreak has the potential to persist for prolonged periods, with the caveat that this estimate was made during the ongoing outbreak.
\end{itemize}

\section{Future modelling}
Future modelling we aim to perform are:
\begin{itemize}
\item An update of previous ODE models of measles outbreak risk in the overall population according the differing vaccine coverage scenarios.
\item Model measles outbreaks with differing scenarios of measles importation into various population groups based on current introduction rates.
\end{itemize}

\section {Summary of key findings}
\begin{itemize}
\item New Zealand is at risk of frequent measles importation due to travel and endemic measles elsewhere in the world.
\item The cost of the current measles outbreak is estimated to be at least \$750,000.
\item Analyses of outbreak data suggest that measles $R_v$ values often include 1 and in this year, 2014, are well above one. This analysis suggests improved vaccination is a requisite to prevent measles becoming endemic again.
\end{itemize}

@@

\bibitem[Carabin et al. (2003)]{carabin3}
NA
\newblock The cost of measles in industrialised countries.
\newblock \emph{Vaccine}, 21,4167--4177.

@@

\bibitem[Dayan et al.(2005)]{dayan5}
NA
\newblock The cost of containing one case of measles: the economic impact on the public health infrastructure - Iowa, 2004.
\newblock \emph{Pediatrics}, 116:e1; DOI:10/1542/peds.2004-2512.

\bibitem[Diekmann et al.(2000)]{diekmann0}
Diekmann, O. and  J.~A.~P. Heesterbeek (2000).
\newblock \emph{Mathematical epidemiology of infectious diseases: model building, analysis and interpretation}.
Chichester: Wiley.

\bibitem[Edmunds et al.(2000)]{edmunds0}
Edmunds, W.~J., N.~J. Gay, M. Kretzschmar, R.~G. Pebody and H. Wachman (2000).
\newblock The pre-vaccination epidemiology of measles, mumps and rubella in Europe: implications for modelling studies.
\newblock \emph{Epidemiology and Infection}, 125, 635--650.

\bibitem[Filia et al.(2007)]{filia7}
NA
\newblock Health burden and economic impact of measles-related hospitalization in Italy, 2002-2003.
\newblock \emph{BMC Public Health}, 7,169

@@

\bibitem[Gay et al.(1998)]{gay98}
Gay, N.~J., L. Pelletier, and P. Duclos (1998).
\newblock Modelling the incidence of measles in Canada: an assessment of the options for vaccination policy.
\newblock \emph{Vaccine}, 16, 794--801.

@@

\bibitem[Koopmanschap(1998)]{koopmanschap98}
Koopmanschap, M.~A. (1998).
\newblock Cost-of-illness studies: useful for health policy?
\newblock \emph{Pharmacoeonomics}, 14, 143--148.

@@

\bibitem[Obidia et al.(2012)]{obidia12}
NA
\newblock The R0 package: a toolbox to estimate reproduction numbers for epidemic outbreaks.
\newblock \emph{BMC Medical Informatics and Decision Making}, 2012, 12--147.

\bibitem[Parker et al.(2006)]{parker6}
NA
\newblock Implications of a 2005 measles outbreak in Indiana for sustained elimination of measles in the United States.
\newblock \emph{The New England Journal of Medicine}, 355, 447--455.

@@

\bibitem[Wallinga and Teunis(2004)]{wallinga4}
Wallinga, J., and P. Teunis (2004).
\newblock Different Epidemic Curves for Severe Acute Respiratory Syndrome Reveal Similar Impacts of Control Measures.
\newblock \emph{American Journal of Epidemiology}, 160, 509.

@@

\bibitem[WHO (2013)]{who13}
World Health Organisation measles media centre, January (2013)
\newblock Geneva: World Health Organization.
\href{http://www.who.int/mediacentre/news/notes/2013/measles_20130117/en/}{www.who.int}, accessed July 1, 2014.
